\Title

### 2023-01-31 Grimaldi-etal-BiolCybernetics.tex
intro

\newcommand{\seeFig}[1]{see Fig.~\ref{fig:#1}}%{see Figure~\ref{fig:#1}}
%% The preceding line is only needed to identify funding in the first footnote. If that is unneeded, please comment it out.
%\usepackage{cite}
%\usepackage[usenames,dvipsnames]{xcolor} %The package starts from the basic facilities of the color package, and provides easy driver-independent access to several kinds of color tints, shades, tones, and mixes of arbitrary colors.
%\usepackage{tcolorbox} % define boxes that can be customed, has to be loaded after xcolor package
\usepackage{amsmath,amssymb,amsfonts} % amsmath and amssymb packages, useful for mathematical formulas and symbols
%%\usepackage{float} %allows me to use [H] with figures to insert figures in multicol env (poster)
%\usepackage{ulem} %The package provides an \ul (underline) command which will break over line ends; this technique may be used to replace \em (both in that form and as the \emph command), so as to make output look as if it comes from a typewrite. Caution: with this package, bibliography will make unbreakable and underlined journal titles
%\usepackage[switch]{lineno} % add line numbers
%\modulolinenumbers[5]
\usepackage{physics} %The  goal  of  this  package  is  to  make  typesetting  equations  for  physics  simpler,  faster,  and  more  human-readable.
%\usepackage{lineno} %Adds line numbers to selected paragraphs with reference possible through the LaTeX \ref and \pageref cross reference mechanism.
%\usepackage{hyperref} %The hyperref package is used to handle cross-referencing commands in LaTeX to produce hypertext links in the document
%\hypersetup{
%    colorlinks = false,
%    hidelinks
%    %linkbordercolor = {white},
%}
%\usepackage{array} %An extended implementation of the array and tabular environments which extends the options for column formats, and provides “programmable” format specifications.
%\usepackage{multicol} %Multicol defines a multicols environment which typesets text in multiple columns
%%\usepackage{wrapfig} %Allows figures or tables to have text wrapped around them. wrapfig is used when calling multicols
% \usepackage[position=top]{subfig} %The package provides support for the manipulation and reference of small or ‘sub’ figures and tables within a single figure or table environment. Linked to subfloat.
% for units
\usepackage{siunitx}%The siunitx package provides a  set  of  tools  for  authors  to  typeset  quantities  in  aconsistent  way.
\AtBeginDocument{\RenewCommandCopy\qty\SI}
\DeclareMathOperator*{\argmax}{arg\,max}
\DeclareMathOperator*{\argmin}{arg\,min}

@@

In a review paper, we report experimental evidence for the presence of precise spiking motifs embedded in recordings from biological neural tissues~\cite{grimaldi_precise_2022}. These spatiotemporal patterns of neural activity can be a useful representation used by neurons to perform the computations needed to solve different cognitive tasks. In 1982, Abeles asked if the role of cortical neurons is whether to integrate synaptic inputs or rather to detect coincidences in temporal spiking motifs~\citep{abeles_role_1982}. While the first possibility favors the rate coding theory, the second highlights the function of temporal precision in the neural code. Since, numerous studies demonstrated the emergence of synchronicity in the activity within a neural population~\citep{riehle_spike_1997, davis_spontaneous_2021}, efficient encoding thanks to the use of spike latencies~\citep{perrinet_coding_2004, gollisch_rapid_2008} or precise timing in the auditory system~\citep{deweese_binary_2002, carr_circuit_1990}. All these findings, and more~\citep{bohte_evidence_2004}, highlight the importance of the temporal aspect of the neural code and suggest the existence of spatio-temporal spiking motifs in biological spike trains. In neuronal models, an efficient use or detection of these spatio-temporal motifs embedded in the spike train comes with the integration of heterogeneous delays~\citep{guise_bayesian_2014, zhang_supervised_2020, nadafian}. Notably,~\citet{izhikevich_polychronization_2006} introduced the notion of the polychronous group as a repetitive motif of spikes defined by a subset of neurons with different, yet precise, relative spiking delays. This representation has a much greater information capacity in comparison to a firing-rate based neural coding approach through the variety of configurations and the possible coexistence of multiple superposed motifs.

The present paper proposes additional experiments on a recent model of spiking neurons with heterogeneous synaptic delays (i.e. time between the arrival of an afferent spike at a specific synapse and the contribution of the associated postsynaptic potential to the neuron's soma)~\cite{}. This model was trained to solve a motion detection task on a synthetic event-based dataset. Once trained, its event-driven computations could be reduced drastically through the pruning of synapses, while maintaining top performance for classification. This showed that we may use the precise timing of spikes to enhance neural computations. In the different kernels obtained after training, one could observe spiking motifs linked to the different motion directions

A new generation of algorithms has emerged with the advent of so-called silicon retinas, which mimics the output of the retina by representing visual information using spikes. %A novel emerging representation is that provided by event-based cameras, in which each pixel independently processes its input and emits an event for positive or negative increments of the log-luminance. 
The shift from the classical, dense representations to this sparse encoding of visual information offers a better analogy to neurobiology, but also offers more energy-efficient computations. It also allows using the precise timing of events. In the HOTS model for instance, the event stream induces spatio-temporal relationships between events which are represented by building ``time surfaces'', 2D images computed using the time difference to the last recorded events~\citep{lagorce_hots_2017}. In a recent study, we have introduced a SNN as a classification model applied to event-based streams using Multinomial Logistic Regression (MLR)~\citep{grimaldi_robust_2022} which extended the HOTS model. By transforming each event in the stream as a vectorial input, our MLR classifier was able to make a decision for every single event, that is, in an online fashion. We have demonstrated on several benchmark datasets that it provides with efficient computations, resulting in ultra-fast categorization. Additionally, we made a formal bridge between this event-based MLR and a SNN, demonstrating the bio-plausibility of this method and its possible integration to neuromorphic hardware. A serie of similar algorithms are applied to real-life applications such as optical flow~\citep{benosman_asynchronous_2012}, depth estimation~\citep{dardelet_event-by-event_2021}, or gesture recognition~\citep{maro_event-based_2020}. Recently, such an algorithm was used in the CASPR mission on the International Space Station, illustrating, given the stringent selection processes used by the space sector, that such algorithms have reached maturity for industry.

This paper explores the possibility that prototypical spiking motifs may be a useful representation used in neurobiology and that we may transfer this hypothesis to neuromorphic algorithms, notably by using Heterogeneous Delays in Spiking Neural Networks (HD-SNN). Our HD-SNN model starts as a simple layer of leaky integrate-and-fire neurons whose connections are defined by a matrix of synaptic weights. Here, we propose to extend our model to a layer of spiking neurons which include heterogeneous delays in addition to weights. In particular, one afferent may be connected with multiple delays. Crucially, we will explicitly use the delay in the computational process. As a consequence, this extends our previous MLR model~\citep{grimaldi_robust_2022} by explictly encoding spiking motifs as spatio-temporal patterns. The objective in this model, by including the dimension of temporal delays, is to increase the representational capacities of the classifier. In the perspective of building energy-efficient algorithms, we will also titrate quantitatively the best trade-off between robustness and computation time when increasing the number of these heterogeneous delays. %

@@

In this work, we study the emergence of spatio-temporal spiking motifs when training a single layer of spiking neurons on a semi-supervised classification task (\seeFig{motion_task}). The paper is organized as follows. We develop a theoretically defined HD-SNN capable of learning heterogeneous delays that we first evaluate on a synthetic event-based dataset. We first detail the methodology by defining the basic mechanism of spiking neurons that utilize heterogeneous delays. This will allow us to formalize the spiking neuron used to learn the model's parameters and test its effectiveness. We aim to apply this model to a realistic motion detection task which we will first define. Then, we will show how the model can be adapted to solve the motion detection task. In the results section, we will detail the emergent structure of the model solving the task and show in particular the similarities with neurobiological observations. We will also test the network with classical stimuli used in neuroscience, and in particular with frequency or contrast responses.  We will also study the efficiency of the motion detection mechanism and in particular its resilience to synaptic weight pruning. This will allow us to show how such a model can provide an efficient solution to the energy/accuracy trade-off.  Finally, we will conclude by highlighting the main contributions of this paper, while defining some limitations. This will open perspectives for future SNNs.  In particular, as neuromorphic devices are by design good candidates for integrating computations over time, we highlight the fact that this event-driven algorithm is perfectly fit to be transfered to this type of hardware and to obtain significant gains in the energy which is used.

>>>>

\newcommand{\seeFig}[1]{see Fig.~\ref{fig:#1}}%{see Figure~\ref{fig:#1}}
%% The preceding line is only needed to identify funding in the first footnote. If that is unneeded, please comment it out.
%\usepackage{cite}
%\usepackage[usenames,dvipsnames]{xcolor} %The package starts from the basic facilities of the color package, and provides easy driver-independent access to several kinds of color tints, shades, tones, and mixes of arbitrary colors.
%\usepackage{tcolorbox} % define boxes that can be customed, has to be loaded after xcolor package
\usepackage{amsmath,amssymb,amsfonts} % amsmath and amssymb packages, useful for mathematical formulas and symbols
%%\usepackage{float} %allows me to use [H] with figures to insert figures in multicol env (poster)
%\usepackage{ulem} %The package provides an \ul (underline) command which will break over line ends; this technique may be used to replace \em (both in that form and as the \emph command), so as to make output look as if it comes from a typewrite. Caution: with this package, bibliography will make unbreakable and underlined journal titles
%\usepackage[switch]{lineno} % add line numbers
%\modulolinenumbers[5]
\usepackage{physics} %The  goal  of  this  package  is  to  make  typesetting  equations  for  physics  simpler,  faster,  and  more  human-readable.
%\usepackage{lineno} %Adds line numbers to selected paragraphs with reference possible through the LaTeX \ref and \pageref cross reference mechanism.
%\usepackage{hyperref} %The hyperref package is used to handle cross-referencing commands in LaTeX to produce hypertext links in the document
%\hypersetup{
%    colorlinks = false,
%    hidelinks
%    %linkbordercolor = {white},
%}
%\usepackage{array} %An extended implementation of the array and tabular environments which extends the options for column formats, and provides “programmable” format specifications.
%\usepackage{multicol} %Multicol defines a multicols environment which typesets text in multiple columns
%%\usepackage{wrapfig} %Allows figures or tables to have text wrapped around them. wrapfig is used when calling multicols
% \usepackage[position=top]{subfig} %The package provides support for the manipulation and reference of small or ‘sub’ figures and tables within a single figure or table environment. Linked to subfloat.
% for units
\usepackage{siunitx}%The siunitx package provides a  set  of  tools  for  authors  to  typeset  quantities  in  aconsistent  way.
%\AtBeginDocument{\RenewCommandCopy\qty\SI}
\DeclareMathOperator*{\argmax}{arg\,max}
\DeclareMathOperator*{\argmin}{arg\,min}

@@

In a review paper, we report experimental evidence for the presence of precise spiking motifs embedded in recordings from biological neural tissues~\cite{grimaldi_precise_2022}. These spatiotemporal patterns of neural activity can be a useful representation used by neurons to perform the computations needed to solve different cognitive tasks. In 1982, Abeles asked if the role of cortical neurons is whether to integrate synaptic inputs or rather to detect coincidences in temporal spiking motifs~\citep{abeles_role_1982}. While the first possibility favors the rate coding theory, the second highlights the function of temporal precision in the neural code. Since, numerous studies demonstrated the emergence of synchronicity in the activity within a neural population~\citep{riehle_spike_1997, davis_spontaneous_2021}, efficient encoding thanks to the use of spike latencies~\citep{perrinet_coding_2004, gollisch_rapid_2008} or precise timing in the auditory system~\citep{deweese_binary_2002, carr_circuit_1990}. All these findings, and more~\citep{bohte_evidence_2004}, highlight the importance of the temporal aspect of the neural code and suggest the existence of spatio-temporal spiking motifs in biological spike trains. In neuronal models, an efficient use or detection of these spatio-temporal motifs embedded in the spike train comes with the integration of heterogeneous delays~\citep{guise_bayesian_2014, zhang_supervised_2020, nadafian_bio-plausible_2020}. Notably,~\citet{izhikevich_polychronization_2006} introduced the notion of the polychronous group as a repetitive motif of spikes defined by a subset of neurons with different, yet precise, relative spiking delays. This representation has a much greater information capacity in comparison to a firing-rate based neural coding approach through the variety of configurations and the possible coexistence of multiple superposed motifs.

The present paper proposes additional experiments on a recent model of spiking neurons with heterogeneous synaptic delays (i.e. time between the arrival of an afferent spike at a specific synapse and the contribution of the associated postsynaptic potential to the neuron's soma)~\cite{grimaldi_learning_2022}. This model was trained to solve a motion detection task on a synthetic event-based dataset. Once trained, its event-driven computations could be reduced drastically through the pruning of synapses, while maintaining top performance for classification. This showed that formal neurons can make use the precise timing of spikes to enhance neural computations. 
In the present paper, we extend the motion detection to a more ecological task. Instead of synthetic textures that were used to generate stimuli for which motion is controlled, we use natural images and random movements that mimic biological saccades. 

@@

Through an ecological motion detection task, we study the emergence of spatio-temporal spiking motifs when training a single layer of spiking neurons on a semi-supervised classification task (\seeFig{motion_task}). First, we develop a theoretically defined Heterogeneous Delays in Spiking Neural Network (HD-SNN) model capable of learning heterogenous synpatic delays. Then, we define the ecological cognitive task the model has to solve with the different datasets it will be tested on. We will study the efficiency of the motion detection mechanism and in particular its resilience to synaptic weight pruning. This will allow us to show how such a model can provide an efficient solution to the energy/accuracy trade-off. 
